---
title: "Project1"
format: html
editor: visual
---

# Project 1

## Executive Summary:

## Prerequisites:

```{r}
#| message: false
library(httr)
library(jsonlite)
library(tidyverse)
apiKey = "4c8e34087cd2fa65ceac20aca5e806b356fffe37"
```

## Working with the PUMS API

First, we will test the API to see the types of data that it provides

```{r}

resp = httr::GET("https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=region:*&SCHL=24&key=4c8e34087cd2fa65ceac20aca5e806b356fffe37")

resp$status_code
```

```{r}
respString = rawToChar(resp$content)

respString
```

The data looks to be formatted like a list of lists. The first row are the requested column headers, and each subsequent row represents an individual in the census. We will need to figure out how to translate the the large character string into a tibble.

After some research, this is a form of JSON, so we can use the fromJSON() function to convert to a matrix.

First we will see if we can convert to a matrix, then let's confirm.

```{r}

respMat = (fromJSON((respString)))

class(respMat)
```

This is a matrix, try to format as a tibble

```{r}
census = as_tibble(respMat)

head(census)
```

This is better, but the first row should be column headers.

```{r}



names(census) = census[1,]

census = census |> slice(-1)

head(census)


```

Looks good!

## Helper Function to Convert API Response

The helper function will take a raw response from an API call and return a tibble.

This function makes certain assumptions about how the data from the Census API are structured.

Assumptions:

-   The first row of the returned data will always contain the row names
-   The schema will always be JSON list of lists

```{r}


rawtoTibble = function(rawResp){
  respTibble = as_tibble(fromJSON((respString)))
  names(respTibble) = respTibble[1, ]
  respTibble = respTibble |>  slice(-1)
  
  return(respTibble)
}


```

Test with an API response:

```{r}
test = httr::GET("https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=region:*&SCHL=24&key=4c8e34087cd2fa65ceac20aca5e806b356fffe37")

head(rawtoTibble(test$content))
```

## Write Main Function

This function will take in a number of user inputs and return a tibble with the census data corresponding to the API call

Requirements:

Year:

-   Inputs

    -   Should be a integer between 2010 and 2022

Numeric Variables:
Options for the user should be AGEP, GASP, GRPIP, JWAP (time), JWDP (time), and
JWMNP

Your function should turn variables into numeric values or time values (use the middle of the time period) where appropriate. This might require you to write a helper function or two!

For missing time values, assign the value an NA

One numeric variable other than PWGTP must be returned (Check for this in input)

```{r}
getCensusData = function(year = 2022, 
                         numericVars = c("AGEP"),
                         categoricalVars = c("SEX"),
                         geoLevel = "State",
                         spec = "NC"){
  if (!(year > 2009) & (year < 2023)){
    stop("Bad Year Input. Try Again")
  }
  
  
  
  acceptableNumericVars = c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  for (i in length(numericVars)){
    ##IF  PWGTP is given as an input, remove it from the vector, as we will return it anyway
    if(numericVars[i] == "PWGTP"){
      numericVars = numericVars[-c(i)]
    }
    ## Check to see if the numeric variables provided are in the list
    if (!(numericVars[i] %in% acceptableNumericVars)){
      stop("Bad Numeric Variable Input. Try Again")
    }
  }
  # If we don't request anything, stop
  if (length(numericVars) == 0){
    stop("You must select one of the following numeric variables: 
         AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  ## Categorial
  acceptableCategoricalVars = c("FER", "HHL", "SCH", "SCHL","SEX")
  for (i in length(categoricalVars)){
    ## Check to see if the categorical variables provided are in the list
    if (!(categoricalVars[i] %in% acceptableCategoricalVars)){
      stop("Bad Categorical Variable Input. Try Again")
    }
  }
  
  ## Region: Check if the user only gives 1 region and it's in the list
  acceptableRegionLabels = c("Region", "Division", "State")
  if (!(geoLevel %in% acceptableRegionLabels) & (length(geoLevel) ==1)){
    stop("Bad Geography Variable Input Try Again")
  }
  
  ## Build API Call URL
  
  baseUrl = "https://api.census.gov/data/2024/acs/acs1/pums?"
  
  
  
  
  return("Made it")
}



getCensusData(geoLevel = c("Region"))
```
